QRF Covariate Selection

### 2019-12-27 analysis/R_scripts/fit_qrf_model.Rmd

>>>>

Because most of our habitat data came from CHaMP sites, which were selected through a generalized random tesselation stratified (GRTS) sample design [@Stevens2003], we wanted to incorporate the design weights for each CHaMP site with a predicted carrying capacity.

```{r survey_design_strata}
#----------------------------------------
# pull in survey design related data
#----------------------------------------
# Calculate GRTS design weights.

# pull in info about what strata each CHaMP site was assigned to (using 2014 as reference year)
site_strata = pred_hab_sites_chnk %>%
  select(Site, Watershed) %>%
  distinct() %>%
  left_join(gaa %>%
              select(Site,
                     # CHaMPsheds,
                     strata = AStrat2014)) %>%
  mutate(site_num = str_split(Site, '-', simplify = T)[,2]) %>%
  mutate(strata = if_else(Watershed == 'Asotin',
                          site_num,
                          if_else(Watershed == 'Entiat' & grepl('ENT00001', Site),
                                  paste('EntiatIMW', site_num, sep = '_'),
                                  strata))) %>%
  mutate(strata = if_else(grepl('EntiatIMW', strata),
                          str_remove(strata, '[[:digit:]]$'),
                          strata),
         strata = if_else(grepl('EntiatIMW', strata),
                          str_remove(strata, '[[:digit:]]$'),
                          strata)) %>%
  filter(!is.na(strata)) %>%
  mutate(strata = paste(Watershed, strata, sep = '_')) %>%
  select(-site_num)

# read in data from the CHaMP frame
champ_frame_df = read_csv('../data/derived_data/champ_frame_data.csv') %>%
  mutate(Target2014 = ifelse(is.na(AStrat2014), 'Non-Target', Target2014)) %>%
  mutate(AStrat2014 = ifelse(AStrat2014 == 'Entiat IMW', paste('EntiatIMW', GeoRchIMW, sep = '_'), AStrat2014)) %>%
  filter(Target2014 == 'Target') %>%
  rename(Watershed = CHaMPshed)

# what strata do we have?
frame_strata = champ_frame_df %>%
  mutate(strata = paste(Watershed, AStrat2014, sep='_')) %>%
  select(Watershed, 
         strata) %>%
  distinct()

# how long is each strata?
chnk_strata_length = champ_frame_df %>%
  filter(!is.na(UseTypCHSP)) %>%
  mutate(strata = paste(Watershed, AStrat2014, sep='_')) %>%
  select(Watershed, matches("Strat"), FrameLeng) %>%
  group_by(Watershed, strata) %>%
  summarise(tot_length_km = sum(FrameLeng) / 1000) %>%
  ungroup() %>%
  mutate_at(vars(Watershed, strata), 
            list(as.factor)) %>%
  arrange(Watershed, strata)

sthd_strata_length = champ_frame_df %>%
  filter(!is.na(UseTypSTSU)) %>%
  mutate(strata = paste(Watershed, AStrat2014, sep='_')) %>%
  select(Watershed, matches("Strat"), FrameLeng) %>%
  group_by(Watershed, strata) %>%
  summarise(tot_length_km = sum(FrameLeng) / 1000) %>%
  bind_rows(tibble(Watershed = 'Asotin',
                   strata = paste('Asotin', c('CC', 'NF', 'SF'), sep = '_'),
                   tot_length_km = 12)) %>%
  ungroup() %>%
  mutate_at(vars(Watershed, strata), 
            list(as.factor)) %>%
  arrange(Watershed, strata)

# how many sites in each strata? and what is the length of each strata?
strata_tab = pred_hab_sites_chnk %>%
  select(Site, Watershed, matches('per_m')) %>%
  left_join(site_strata) %>%
  filter(strata != 'Entiat_Entiat IMW') %>%
  mutate_at(vars(Watershed),
            list(fct_drop)) %>%
  group_by(Watershed, strata) %>%
  summarise(n_sites = n_distinct(Site)) %>%
  ungroup() %>%
  full_join(chnk_strata_length) %>%
  mutate(n_sites = if_else(is.na(n_sites),
                           as.integer(0),
                           n_sites)) %>%
  # calculate the weight of each site in each strata
  mutate(site_weight = if_else(n_sites > 0,
                               tot_length_km / n_sites,
                               as.numeric(NA)))

strata_test = frame_strata %>%
  full_join(strata_tab) %>%
  mutate_at(vars(Watershed),
            list(fct_drop)) %>%
  mutate(n_sites = if_else(is.na(n_sites),
                           as.integer(0),
                           n_sites))

# # what frame strata don't have any sites in them?
# strata_test %>%
#   filter(n_sites == 0,
#          !is.na(tot_length_km)) %>%
#   arrange(Watershed, strata) %>%
#   as.data.frame()

# # what strata that we have sites for are not in the frame strata?
# strata_test %>%
#   filter(n_sites > 0,
#          (is.na(tot_length_km) |
#             tot_length_km == 0)) %>%
#   as.data.frame()

# how much of each watershed is not accounted for with current sites / strata?
strata_test %>%
  group_by(Watershed) %>%
  summarise_at(vars(tot_length_km),
               list(sum),
               na.rm = T) %>%
  left_join(strata_test %>%
              filter(n_sites == 0) %>%
              group_by(Watershed) %>%
              summarise_at(vars(missing_length = tot_length_km),
                           list(sum),
                           na.rm = T)) %>%
  mutate_at(vars(missing_length),
            list(~ if_else(is.na(.), 0, .))) %>%
  mutate(perc_missing = missing_length / tot_length_km) %>%
  mutate_at(vars(perc_missing),
            list(~ if_else(is.na(.), 0, .))) %>%
  arrange(desc(perc_missing)) %>%
  pander()

We then selected a subset of the globally available attributes (GAAs) to use as extrapolation covariates. This was based on minimizing the correlations between the covariates. We also acknowlegdged which prediction sites had GAAs outside the range of values for the CHaMP sites.

We use the log of predicted capacity wherever we could make those predictions as the response variable, and this group of selected GAAs as the covariates. We then fit a linear model that incorporated the survey design weights by using the `survey` pacakge (-@Lumley2014) in R. We fit two main models, one that included CHaMP watershed as a covariate, and one that did not. The former was used to make predictions within all CHaMP watersheds where we had QRF predictions of carrying capacity. The latter was used at all other master sample points.

```{r}
data(gaa)

gaa_all = gaa %>%
  rename(Channel_Type = ChanlType)

# possible covariates from among the GAAs
gaa_covars = c('TRange', 'GDD', 'Precip', 'Elev_M', 'CHaMPsheds', 'NatPrin1', 'NatPrin2', 'DistPrin1', 'BFW_M', 'SrtCumDrn', 'StrmPwr', 'Slp_NHD_v1', 'Channel_Type', 'MAVELV', 'WIDE_BF', 'S2_02_11')

# what type of covariate is each GAA?
gaa_class = gaa_all %>%
  select(one_of(gaa_covars)) %>%
  as.list() %>%
  map_chr(.f = class)

# which ones are numeric?
gaa_num = names(gaa_class)[gaa_class %in% c('integer', 'numeric')]
# which ones are categorical?
gaa_catg = names(gaa_class)[gaa_class %in% c('factor', 'character')]

# compare range of covariates from model dataset and prediction dataset
range_comp = bind_rows(gaa_all %>%
                         filter(!Site %in% unique(pred_hab_sites$Site)) %>%
                         select(Site,
                                one_of(gaa_num)) %>%
                         gather(Metric, value, -Site) %>%
                         mutate(Source = 'non-DASH Sites'),
                       gaa_all %>%
                         filter(Site %in% unique(pred_hab_sites$Site)) %>%
                         select(Site,
                                one_of(gaa_num)) %>%
                         distinct() %>%
                         gather(Metric, value, -Site) %>%
                         mutate(Source = 'DASH Sites')) %>%
  mutate_at(vars(Source, Metric),
            list(as.factor))

range_max = range_comp %>%
  group_by(Metric, Source) %>%
  summarise_at(vars(value),
               tibble::lst(min, max),
               na.rm = T) %>%
  filter(Source == 'DASH Sites') %>%
  ungroup() %>%
  gather(type, value, -Metric, -Source)

covar_range_p = range_comp %>%
  ggplot(aes(x = Source,
             y = value,
             fill = Source)) +
  geom_boxplot() +
  facet_wrap(~ Metric,
             scales = 'free') +
  geom_hline(data = range_comp %>%
               group_by(Metric, Source) %>%
               summarise_at(vars(value),
                            tibble::lst(min, max),
                            na.rm = T) %>%
               filter(Source == 'DASH Sites') %>%
               ungroup() %>%
               gather(type, value, min, max),
             aes(yintercept = value),
             lty = 2,
             color = 'darkgray') +
  theme_minimal()

covar_range_p

# look at correlations among GAAs
library(corrr)
corr_df = gaa_all %>%
  select(one_of(gaa_num)) %>%
  correlate()

corr_df %>%
  shave(upper = T) %>%
  stretch() %>%
  arrange(desc(abs(r))) %>%
  filter(abs(r) >= 0.5)

gaa_corr_p1 = corr_df %>%
  rearrange(absolute = T) %>%
  shave(upper = T) %>%
  rplot(legend = T,
        colors = c('blue', 'white', 'indianred2'),
        print_cor = T)

# correlation matrix
corr_mat = gaa_all %>%
  select(one_of(gaa_num)) %>%
  cor(use = 'pairwise.complete.obs')

library(ggcorrplot)
gaa_corr_p2 = ggcorrplot(corr_mat,
           lab = T,
           lab_size = 2,
           hc.order = T,
           colors = c('#2166AC', 'white', '#B2182B'),
           tl.cex = 8,
           type = 'full')
```

```{r}
# reduce number of covariates bases on correlations
gaa_covars = c('TRange', 
               # 'GDD', 
               # 'Precip',
               'Elev_M', 
               'CHaMPsheds', 
               'NatPrin1', 
               # 'NatPrin2', 
               'DistPrin1', 
               # 'BFW_M', 
               'SrtCumDrn', 
               'StrmPwr',
               'Slp_NHD_v1', 
               'Channel_Type', 
               # 'MAVELV', 
               'WIDE_BF',
               'S2_02_11')

# what type of covariate is each GAA?
gaa_class = gaa_all %>%
  select(one_of(gaa_covars)) %>%
  as.list() %>%
  map_chr(.f = class)

# which ones are numeric?
gaa_num = names(gaa_class)[gaa_class %in% c('integer', 'numeric')]
# which ones are categorical?
gaa_catg = names(gaa_class)[gaa_class %in% c('factor', 'character')]

# Center the covariates
# filter out sites with covariates outside range of covariates used to fit extrapolation model
out_range_sites = gaa_all %>%
  # filter out a few areas
  filter(!HUC6NmNRCS %in% c('Upper Sacramento', 'Southern Oregon Coastal', 'Puget Sound', 'Northern California Coastal', 'Oregon Closed Basins')) %>%
  # don't use AEM sites in model
  filter(!grepl('^AEM', Site)) %>%
  select(one_of(gaa_num), Site) %>%
  gather(Metric, value, -Site) %>%
  left_join(select(range_max, -Source) %>%
              spread(type, value)) %>%
  group_by(Metric) %>%
  filter(value > max |
           value < min) %>%
  ungroup() %>%
  pull(Site) %>%
  unique()

# center covariates
gaa_summ = inner_join(pred_hab_sites %>%
                          select(Site) %>%
                          distinct(),
                        gaa_all %>%
                          select(Site, one_of(gaa_num))) %>%
  gather(GAA, value, -Site) %>%
  group_by(GAA) %>%
  summarise(metric_mean = mean(value, na.rm=T),
            metric_sd = sd(value, na.rm=T)) %>%
  ungroup()

# extrapolation model data set, with normalized covariates
mod_data = inner_join(pred_hab_sites_chnk %>%
                        select(Site:avg_aug_temp,
                               matches('per_m'),
                               -one_of(gaa_num)),
                      gaa_all %>%
                        select(Site, one_of(gaa_num))) %>%
  gather(GAA, value, one_of(gaa_num)) %>%
  left_join(gaa_summ) %>%
  mutate(norm_value = (value - metric_mean) / metric_sd) %>%
  select(-(value:metric_sd)) %>%
  spread(GAA, norm_value) %>%
  left_join(gaa_all %>%
              select(Site, one_of(gaa_catg)))

mod_data %<>%
  bind_cols(mod_data %>%
              is.na() %>%
              as_tibble() %>%
              select(one_of(gaa_covars)) %>%
              transmute(n_NA = rowSums(.))) %>%
  filter(n_NA == 0) %>%
  mutate_at(vars(Watershed, CHaMPsheds, Channel_Type),
            list(fct_drop))

# where do we want to make extrapolation predictions?
gaa_pred = gaa_all %>%
  # filter out a few areas
  filter(!HUC6NmNRCS %in% c('Upper Sacramento', 'Southern Oregon Coastal', 'Puget Sound', 'Northern California Coastal', 'Oregon Closed Basins')) %>%
  # don't use AEM sites in model
  filter(!grepl('^AEM', Site)) %>%
  # don't use non-GRTS sites
  filter(SiteID_alt != 'NonGRTSSite' | is.na(SiteID_alt)) %>%
  filter(!grepl('mega', Site, ignore.case = T)) %>%
  # note which sites have GAAs outside range of CHaMP sites GAAs
  mutate(inCovarRange = ifelse(Site %in% out_range_sites, F, T)) %>%
  select(Site, one_of(gaa_covars), Lon, Lat, inCovarRange, HUC6NmNRCS, HUC8NmNRCS, HUC10NmNRC, HUC12NmNRC, chnk) %>%
  gather(GAA, value, one_of(gaa_num)) %>%
  left_join(gaa_summ) %>%
  mutate(norm_value = (value - metric_mean) / metric_sd) %>%
  select(-(value:metric_sd)) %>%
  spread(GAA, norm_value)

```{r with_model_weights, eval = T}
# calculate adjusted weights for all predicted QRF capacity sites
mod_data_weights = mod_data %>%
  left_join(site_strata) %>%
  left_join(strata_tab) %>%
  filter(!is.na(site_weight)) %>%
  group_by(Watershed) %>%
  mutate(sum_weights = sum(site_weight)) %>%
  ungroup() %>%
  mutate(adj_weight = site_weight / sum_weights)

#-------------------------------------------------------------
# Set up the survey design.

# getOption('survey.lonely.psu')
# this will prevent strata with only 1 site from contributing to the variance
# options(survey.lonely.psu = 'certainty')
# this centers strata with only 1 site to the sample grand mean; this is conservative
options(survey.lonely.psu = 'adjust')

# extrapolation model formula
full_form = as.formula(paste('log(qrf_cap) ~ -1 + (', paste(gaa_covars, collapse = ' + '), ')'))

# fit various models
model_svy_df = mod_data_weights %>%
  gather(response, qrf_cap, matches('per_m')) %>%
  select(-(n_sites:sum_weights)) %>%
  group_by(response) %>%
  nest() %>%
  mutate(design = map(data,
                      .f = function(x) {
                        svydesign(id = ~ 1,
                                  data = x,
                                  strata = ~ Watershed,
                                  # strata = ~ strata,
                                  weights = ~ adj_weight)
                      })) %>%
  mutate(mod_full = map(design,
                        .f = function(x) {
                          svyglm(full_form,
                                 design = x)
                        }),
         mod_no_champ = map(design,
                            .f = function(x) {
                              svyglm(update(full_form, .~ . -CHaMPsheds),
                                     design = x)
                            }))
```

```{r make-predictions}
# predictions within CHaMP watersheds, using the model that includes CHaMPsheds as a covariate
y = gaa_pred %>%
  filter(CHaMPsheds %in% unique(model_svy_df$data[[1]]$Watershed)) %>%
  select(Site, CHaMPsheds, one_of(gaa_covars)) %>%
  na.omit() %>%
  left_join(gaa_pred) %>%
  mutate_at(vars(Channel_Type),
            list(as.factor))

per_m_preds = predict(model_svy_df$mod_full[[1]],
                      newdata = y,
                      se = T,
                      type = 'response')

per_m2_preds = predict(model_svy_df$mod_full[[2]],
                      newdata = y,
                      se = T,
                      type = 'response')

y %<>%
  bind_cols(per_m_preds %>%
              as_tibble() %>%
              rename(chnk_per_m = response,
                     chnk_per_m_se = SE)) %>%
  bind_cols(per_m2_preds %>%
              as_tibble() %>%
              rename(chnk_per_m2 = response,
                     chnk_per_m2_se = SE)) %>%
  mutate_at(vars(matches('per_m')),
            list(exp))

rm(per_m_preds, per_m2_preds)

# predictions at all points, using the model without CHaMPsheds as a covariate
z = gaa_pred %>%
  select(Site, one_of(gaa_covars), -CHaMPsheds) %>%
  na.omit() %>%
  left_join(gaa_pred) %>%
  mutate_at(vars(Channel_Type),
            list(as.factor))

per_m_preds = predict(model_svy_df$mod_no_champ[[1]],
                      newdata = z,
                      se = T,
                      type = 'response')

per_m2_preds = predict(model_svy_df$mod_no_champ[[2]],
                      newdata = z,
                      se = T,
                      type = 'response')

z %<>%
  bind_cols(per_m_preds %>%
              as_tibble() %>%
              rename(chnk_per_m = response,
                     chnk_per_m_se = SE)) %>%
  bind_cols(per_m2_preds %>%
              as_tibble() %>%
              rename(chnk_per_m2 = response,
                     chnk_per_m2_se = SE)) %>%
  mutate_at(vars(matches('per_m')),
            list(exp))

rm(per_m_preds, per_m2_preds)

all_preds = y %>%
  mutate(model = 'CHaMP') %>%
  bind_rows(z %>%
              mutate(model = 'non-CHaMP'))

# quick comparison of capacity predicitons with both models
comp_pred_p = all_preds %>%
  filter(Site %in% Site[duplicated(Site)]) %>%
  select(Site, CHaMPsheds, Channel_Type, model, chnk_per_m, chnk_per_m2) %>%
  arrange(CHaMPsheds, Site, model) %>%
  gather(dens_type, cap, starts_with('chnk_per_m')) %>%
  spread(model, cap) %>%
  ggplot(aes(x = CHaMP,
             y = `non-CHaMP`)) +
  geom_point() +
  geom_abline(linetype = 2,
              color = 'red') +
  facet_wrap(~ CHaMPsheds + dens_type,
             scales = 'free')

# for sites in CHaMP watersheds, use predicions from CHaMP extrapolation model
all_preds %<>%
  filter((Site %in% Site[duplicated(Site)] & model == 'CHaMP') |
           !Site %in% Site[duplicated(Site)])

```{r save-extrapolation}

usethis::use_data(gaa_covars,
                  mod_data_weights,
                  model_svy_df,
                  all_preds,
                  version = 2,
                  overwrite = T)
